---
title: 'Computation Immunology HW #1'
author: "Aislinn Gilmour"
date: "2025-09-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Problem 1** 
*Output from first 10 lines of the gene expression file for SRR8281233:*
(salmon) [f006fph@polaris8 SRR8281233]$ head quant.sf
Name	Length	EffectiveLength	TPM	NumReads
ENST00000832824.1	1379	1213.242	0.000000	0.000
ENST00000832825.1	1417	1251.242	0.000000	0.000
ENST00000832826.1	1538	1372.242	0.000000	0.000
ENST00000832827.1	1519	1353.242	0.000000	0.000
ENST00000832828.1	1762	1596.242	0.000000	0.000
ENST00000832829.1	1336	1170.242	0.169722	3.383
ENST00000832830.1	1260	1094.242	0.000000	0.000
ENST00000832837.1	1553	1387.242	0.000000	0.000
ENST00000832836.1	1636	1470.242	0.000000	0.000

*Name*: 
*Length*:
*Effective Length*:
*TPM*:
*NumReads*:

**Problem 2**
*Problem 2.1*: Analyzing with DESeq2 for differential gene expression analysis between responders (R) and non-responders (NR) (20 points)
2.1.1. Install DESeq2. Recommend installing the packages through bioconductor. Additional packages: tximport, EnhancedVolcano
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("DESeq2")
BiocManager::install("tximport")
BiocManager::install('EnhancedVolcano')
BiocManager::install("apeglm") # It's will be used in the lfcShrink after DESeq2
```

Load Libraries
```{r}
library(DESeq2)
library(tximport)
library(EnhancedVolcano)
library(tidyr)
library(readr)
```

2.1.2. Read in Salmon Data
```{r}
data = read.csv("dataset.csv", header=T)
files <- file.path("salmon", data$Sample, "quant.sf")
tx2gene_full <- read.csv("gencode.v48.tx2gene.csv")
tx2gene <- tx2gene_full[c("transcript_id", "gene_name")]
txi <- tximport(files, type="salmon", tx2gene=tx2gene)
```

2.1.3. Run DESeq2 on the tximported data
```{r}
#convert Response col to factor
data$Response<-as.factor(data$Response)

ddsTxi <- DESeqDataSetFromTximport(txi, colData=data, design=~Response) # Fill in your code here. 
dds<-DESeq(ddsTxi)
dds_res<-results(dds, contrast=c("Response","R","NR")) # Fill in the code here to make sure the comparison order is R vs NR, not NR vs R.
summary(dds_res)
```

2.1.4. Volcano plot on the raw DESeq2 result (log2foldchange and adjusted p value)
```{r}
EnhancedVolcano(dds_res, lab=rownames(dds_res), x="log2FoldChange", y="padj", pCutoff=0.1)
```

2.1.5. Shrink the log2foldchange then examine the volcano plot
```{r}
resultsNames(dds)
dds_resLFCShrink<-lfcShrink(dds, coef="Response_R_vs_NR", res=dds_res) 
EnhancedVolcano(dds_resLFCShrink, lab=rownames(dds_resLFCShrink), x="log2FoldChange", y="padj", pCutoff=0.1)
```
*Question*: Does the lfcShrink change the volcano plot? If so, what changes?

*Problem 2.2*: GO Enrichment Analysis
```{r}
# Load clusterProfilter and the human GO database
# BiocManager::install("clusterProfiler")
# BiocManager::install("org.Hs.eg.db")

library(clusterProfiler)
library(org.Hs.eg.db)
```

```{r}
dds_resLFCShrink$gene_name <- rownames(dds_resLFCShrink)
up_genes <- unique(na.omit(subset(dds_resLFCShrink, padj<0.1 & log2FoldChange > 0))$gene_name) # Use the genes with adjusted p value less than 0.1

go_res<-enrichGO(gene = up_genes,
         OrgDb         = org.Hs.eg.db,
         keyType       = "SYMBOL",
         ont           = "BP",
         pAdjustMethod = "BH",
         qvalueCutoff  = 0.1)
summary(go_res)
```

*Question*: did you find any enrichment GO term? If not, what part of the given code snippet (after filling the missing part) would you change?

*Problem 2.3*: GSEA using clusterProfilter (10 points)
```{r}
#BiocManager::install("enrichplot")

library(enrichplot)
```

```{r}
# ??vs dds_resFCShrink$log2FoldChange
gene_list <- dds_res$stat # Fill your code here. Which feature do you want to use to order the genes?
names(gene_list) <- rownames(dds_res)
gene_list<-na.omit(gene_list)
gene_list <- sort(gene_list, decreasing=TRUE)

gsea_go_res<-gseGO(gene = gene_list,
         OrgDb         = org.Hs.eg.db,
         keyType       = "SYMBOL",
         ont           = "BP",
         pAdjustMethod = "BH")
head(gsea_go_res)
```


```{r}
# Draw the GSEA plot for the most significant pathway 
gseaplot2(gsea_go_res, geneSetID = "GO:0048706")
```








