---
title: 'Computation Immunology HW #1'
author: "Aislinn Gilmour"
date: "2025-09-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Problem 1** 

*Question*: Show the first few lines of the gene expression file (quant.sf) for SRR8281233, and explain what each column represents.
(salmon) [f006fph@polaris8 SRR8281233]$ head quant.sf
Name	Length	EffectiveLength	TPM	NumReads
ENST00000832824.1	1379	1213.242	0.000000	0.000
ENST00000832825.1	1417	1251.242	0.000000	0.000
ENST00000832826.1	1538	1372.242	0.000000	0.000
ENST00000832827.1	1519	1353.242	0.000000	0.000
ENST00000832828.1	1762	1596.242	0.000000	0.000
ENST00000832829.1	1336	1170.242	0.169722	3.383
ENST00000832830.1	1260	1094.242	0.000000	0.000
ENST00000832837.1	1553	1387.242	0.000000	0.000
ENST00000832836.1	1636	1470.242	0.000000	0.000

*Name*: Transcript ID, can be translated to a readable gene name
*Length*: Length of the full underlying transcript
*Effective Length*: Length of the transcript after adjusting for areas of the transcript with low transcript integrity to only keep the usable portions of the transcript for generating fragments
*TPM*: Normalized transcript abundance
*NumReads*: Sum of the weighted read distribution

**Problem 2**
*Problem 2.1*: Analyzing with DESeq2 for differential gene expression analysis between responders (R) and non-responders (NR) (20 points)
2.1.1. Install DESeq2. Recommend installing the packages through bioconductor. Additional packages: tximport, EnhancedVolcano
```{r}
# if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# BiocManager::install("DESeq2")
# BiocManager::install("tximport")
# BiocManager::install('EnhancedVolcano')
# BiocManager::install("apeglm") # It's will be used in the lfcShrink after DESeq2
```

Load Libraries
```{r}
library(DESeq2)
library(tximport)
library(EnhancedVolcano)
library(tidyr)
library(dplyr)
library(readr)
```

2.1.2. Read in Salmon Data
```{r}
data = read.csv("dataset.csv", header=T)
files <- file.path("salmon", data$Sample, "quant.sf")
tx2gene_full <- read.csv("gencode.v48.tx2gene.csv")
tx2gene <- tx2gene_full[c("transcript_id", "gene_name")]
txi <- tximport(files, type="salmon", tx2gene=tx2gene)
```

2.1.3. Run DESeq2 on the tximported data
```{r}
#convert Response col to factor
data$Response<-as.factor(data$Response)

ddsTxi <- DESeqDataSetFromTximport(txi, colData=data, design=~Response) # Fill in your code here. 
dds<-DESeq(ddsTxi)
dds_res<-results(dds, contrast=c("Response","R","NR")) # Fill in the code here to make sure the comparison order is R vs NR, not NR vs R.
summary(dds_res)
```

2.1.4. Volcano plot on the raw DESeq2 result (log2foldchange and adjusted p value)
```{r}
EnhancedVolcano(dds_res, lab=rownames(dds_res), x="log2FoldChange", y="padj", pCutoff=0.1)
```

2.1.5. Shrink the log2foldchange then examine the volcano plot
```{r}
resultsNames(dds)
dds_resLFCShrink<-lfcShrink(dds, coef="Response_R_vs_NR", res=dds_res) 
EnhancedVolcano(dds_resLFCShrink, lab=rownames(dds_resLFCShrink), x="log2FoldChange", y="padj", pCutoff=0.1)
```
*Question*: Does the lfcShrink change the volcano plot? If so, what changes?
Yes, the lfcshrink decreased the noise in the plot and reduced outliers with extreme fold changes so that more genes were pulled to 0 on the X axis and the axis scale was able to shrink. This also allowed for additional genes to come through in the model, inlcuding GOLGA8O and TFAP2A. 

*Problem 2.2*: GO Enrichment Analysis
```{r}
# Load clusterProfilter and the human GO database
# BiocManager::install("clusterProfiler")
# BiocManager::install("org.Hs.eg.db")

library(clusterProfiler)
library(org.Hs.eg.db)
```

```{r}
dds_resLFCShrink$gene_name <- rownames(dds_resLFCShrink)
up_genes <- unique(na.omit(subset(dds_resLFCShrink, padj<0.1 & log2FoldChange > 0))$gene_name) # Use the genes with adjusted p value less than 0.1

go_res<-enrichGO(gene = up_genes,
         OrgDb         = org.Hs.eg.db,
         keyType       = "SYMBOL",
         ont           = "BP",
         pAdjustMethod = "BH",
         qvalueCutoff  = 0.1)
summary(go_res)
```

*Question*: did you find any enrichment GO term? If not, what part of the given code snippet (after filling the missing part) would you change?
No, 

*Problem 2.3*: GSEA using clusterProfilter (10 points)
```{r}
#BiocManager::install("enrichplot")

library(enrichplot)
```

```{r}
# ??vs dds_resFCShrink$log2FoldChange
gene_list <- dds_res$stat # Fill your code here. Which feature do you want to use to order the genes?
names(gene_list) <- rownames(dds_res)
gene_list<-na.omit(gene_list)
gene_list <- sort(gene_list, decreasing=TRUE)

gsea_go_res<-gseGO(gene = gene_list,
         OrgDb         = org.Hs.eg.db,
         keyType       = "SYMBOL",
         ont           = "BP",
         pAdjustMethod = "BH")
head(gsea_go_res)
```


```{r}
# Draw the GSEA plot for the most significant pathway 
gseaplot2(gsea_go_res, geneSetID = "GO:0048706")
```

**Problem 3** Differential gene expression analysis considering batch effect (20 points)

```{r}
library(ggplot2)

# PCA for 2000 genes with highest variances for batch effect inspection
log2tpm <- log2(txi$abundance+1)
rv <- apply(log2tpm, 1, var)
cutoff <- sort(rv, decreasing = TRUE)[2000]
log2tpm_variable <- log2tpm[rv >= cutoff, ]
sample_data_pca <- data.frame(data,prcomp(t(log2tpm_variable),scale = TRUE, center = TRUE)$x)
ggplot2::ggplot(sample_data_pca, ggplot2::aes(x = PC1, y = PC2, shape=Response, color = as.factor(Batch))) +ggplot2::geom_point(size=10)
```

Run DESeq2 on the tximported data- with batch effect in the model
```{r}
#convert Response col to factor
data$Response<-as.factor(data$Response)

ddsTxi <- DESeqDataSetFromTximport(txi, colData=data, design=~Response + as.factor(Batch)) # Fill in your code here. 
dds<-DESeq(ddsTxi)
dds_res<-results(dds, contrast=c("Response","R","NR")) # Fill in the code here to make sure the comparison order is R vs NR, not NR vs R.
summary(dds_res)
```

Volcano plot on the raw DESeq2 result (log2foldchange and adjusted p value)
```{r}
EnhancedVolcano(dds_res, lab=rownames(dds_res), x="log2FoldChange", y="padj", pCutoff=0.1)
```

Shrink the log2foldchange then examine the volcano plot
```{r}
resultsNames(dds)
dds_resLFCShrink<-lfcShrink(dds, coef="Response_R_vs_NR", res=dds_res) 
EnhancedVolcano(dds_resLFCShrink, lab=rownames(dds_resLFCShrink), x="log2FoldChange", y="padj", pCutoff=0.1)
```

GO Enrichment Analysis
```{r}
# Load clusterProfilter and the human GO database
# BiocManager::install("clusterProfiler")
# BiocManager::install("org.Hs.eg.db")

# library(clusterProfiler)
# library(org.Hs.eg.db)
```

```{r}
dds_resLFCShrink$gene_name <- rownames(dds_resLFCShrink)
up_genes <- unique(na.omit(subset(dds_resLFCShrink, padj<0.1 & log2FoldChange > 0))$gene_name) # Use the genes with adjusted p value less than 0.1

go_res<-enrichGO(gene = up_genes,
         OrgDb         = org.Hs.eg.db,
         keyType       = "SYMBOL",
         ont           = "BP",
         pAdjustMethod = "BH",
         qvalueCutoff  = 0.1)
summary(go_res)
```


GSEA using clusterProfilter 
```{r}
#BiocManager::install("enrichplot")

# library(enrichplot)
```

```{r}
# ??vs dds_resFCShrink$log2FoldChange
gene_list <- dds_res$stat # Fill your code here. Which feature do you want to use to order the genes?
names(gene_list) <- rownames(dds_res)
gene_list<-na.omit(gene_list)
gene_list <- sort(gene_list, decreasing=TRUE)

gsea_go_res<-gseGO(gene = gene_list,
         OrgDb         = org.Hs.eg.db,
         keyType       = "SYMBOL",
         ont           = "BP",
         pAdjustMethod = "BH")
head(gsea_go_res)
```


```{r}
# Draw the GSEA plot for the most significant pathway 
gseaplot2(gsea_go_res, geneSetID = "GO:0033108")
```
*Question*: do you think the new results that consider batch effects give more biological meaningful results? What is your biological conclusion from the analysis?

**Problem 4.** Use EPIC for immune cell deconvolution (20 points)
```{r}
# install.packages("devtools")
# devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE) # https://github.com/GfellerLab/EPIC

library(EPIC)
```

```{r}
abundance<- txi$abundance
epic_res<-EPIC(abundance, reference="BRef") # Use blood as reference. 
epic_res$cellFractions
```

*Question*: Visualize the cellFraction difference between R and NR for all the EPIC cell types and calculate the p-value for the difference using Wilcoxon rank-sum test. For example, the fraction of B cells in R vs NR. You may use the code below or implement your own code.

```{r}
library(ggpubr)
# This is an example using ggplot and ggpubr . You may need to install corresponding packages if you decide to use this code.
epic_wmeta <- bind_cols(data, epic_res$cellFractions)

epic_wmeta_long <- epic_wmeta %>% pivot_longer(cols=all_of(colnames(epic_res$cellFractions)),
                                              names_to="celltype", values_to="fraction")

p <- ggplot(epic_wmeta_long, aes(x = Response, y = fraction, fill = Response)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +   
  geom_jitter(width = 0.15, size = 1.2, alpha = 0.6) +
  facet_wrap(~ celltype, scales = "free_y", ncol = 4) + 
  theme_bw() +
  labs(x = "", y = "Cell fraction") +
  theme(legend.position = "none")

p <- p + stat_compare_means(method = "wilcox.test", 
        aes(label=paste0("p=", after_stat(p.format))))

p
```

*Question*: For the cell type with p-value less than 0.2, provide the potential explanation on why this is between R and NR. If you don't see such cell type, then disregard the significance, select one of the cell type to provide the potential difference regardless of p-value.

CD4 T cells have a p-value of 0.1, meaning they are significantly more expressed in non-responders to the immunotherapy for glioblastoma than the responders. This could be because when CD4+ treg cells are overexpressed they are immunosupressive, making the treatment ineffective.



