---
title: "HW 2"
author: "Aislinn Gilmour"
date: "2025-10-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load Libraries
```{r, message=FALSE, warning=FALSE}
library(tidyr)
library(dplyr)
library(Seurat)
library(ggplot2)
library(clustree)
library(harmony)
```

**Get ready for data integration. 20 points **

**a.	Read in all data samples to create a combined object.**

Load in Data from Geo
```{r , message=FALSE, warning=FALSE}
#untar data from GEO
untar("GSE155224_RAW.tar", exdir = "geo_data")

#unzip the cell ranger files
#NOTE- i used Chat GPT to help write this loop because I couldn't get the data to 
#load in correctly and I couldn't find any examples online doing it in a loop 
for (f in list.files("geo_data", pattern = "cellranger\\.tar\\.gz$", full.names = TRUE,
                     recursive = TRUE)) {
  outdir <- file.path("geo_data", "unzipped", sub("\\.tar.*$", "", basename(f)))
  dir.create(outdir, showWarnings = FALSE, recursive = TRUE)
  untar(f, exdir = outdir)
}
```

Get Files and Sample Names
```{r, message=FALSE, warning=FALSE}
# Note: I also used Chat GPT to use the right RegEx for this section to reference the correct file paths

files <- list.files(
  "geo_data/unzipped",
  pattern = "filtered_feature_bc_matrix$",
  recursive = TRUE,
  full.names = TRUE,
  include.dirs = TRUE
)

# Skip CITE files
files <- files[!grepl("CITE", files, ignore.case = TRUE)]

inner   <- basename(dirname(dirname(files)))
samples <- sub("_.*$", "", inner)

```

Load in Data
```{r, message=FALSE, warning=FALSE}
for(i in 1:length(samples)){
 sample = samples[i]
 file = files[i]
 print(sample)
 td = Read10X(file)
 t = CreateSeuratObject(counts = td, project=sample)
 if(sample == samples[1]){
 obj = t
 }else{
 obj = merge(obj, t)
 }
}
```

Set up Patient Cohorts
```{r , message=FALSE, warning=FALSE}
obj@meta.data$cohort = "COVID-19"
obj@meta.data$cohort[nchar(obj@meta.data$orig.ident) == 10] = "Healthy"
```

**b.	Remove cells with >10% mtDNA and < 400 features**
```{r, message=FALSE, warning=FALSE}
# remove cells with < 400 features or >10% mtDNA
p = PercentageFeatureSet(object = obj, pattern = "^MT-")
obj[["percent.mt"]] = p
obj = subset(x = obj, subset = nFeature_RNA > 400 & percent.mt < 10)

# save intermediate file
saveRDS(obj, "~/Documents/Dartmouth Class Work/Fall 2025/QBS145/obj.rds")
```

**c.	Normalize the count data log-transformation**
```{r, message=FALSE, warning=FALSE}
# normalize
obj <- NormalizeData(obj)
```

**d.	Find the top 2000 most variable genes.**
```{r, message=FALSE, warning=FALSE}
# 2000 most variable genes
obj <- FindVariableFeatures(obj)
```

**e.	Remove immunoglobulin (BCR) and TCR-related genes from variable genes list.**
```{r, message=FALSE, warning=FALSE}
# look at variable genes from Ig and TR variable and constant region segments
head(VariableFeatures(obj)[grepl("^IG|^TR",VariableFeatures(obj))])

# Get list of Ig and TR variable and constant region segments
load("QC_features_meta.RData")
biotypes_excl = unique(features_meta[["gene_biotype"]])[grepl(pattern="^IG_|^TR_", x=unique(features_meta[["gene_biotype"]]))]
remove.genes = features_meta[["external_gene_name"]][features_meta[["gene_biotype"]] %in% biotypes_excl]

# remove IG and TR genes from variable genes list
current_variable_features = VariableFeatures(obj)
VariableFeatures(obj) = current_variable_features[!current_variable_features %in% remove.genes]
```


**f. Scale data**
```{r, message=FALSE, warning=FALSE}
# mean 0, variance 1
obj <- ScaleData(obj)
```

**2.	Run PCA, then integrate all samples together using Harmony. 10 points**

PCA
```{r, message=FALSE, warning=FALSE}
obj <- RunPCA(obj, features = VariableFeatures(object = obj))

# print
print(obj[["pca"]], dims = 1:5, nfeatures = 5)
DimPlot(obj, reduction = "pca") + NoLegend()
DimHeatmap(obj, dims = 1, cells = 500, balanced = TRUE)
```


Integrate with Harmony 
```{r, message=FALSE, warning=FALSE}
# Integrate using Harmony
obj.int <- IntegrateLayers(
  object = obj, method = HarmonyIntegration,
  orig.reduction = "pca", new.reduction = "harmony",
  verbose = FALSE
  )
# Make UMAP
obj.int <- RunUMAP(obj.int, reduction = "harmony",
  dims = 1:30, reduction.name = "umap.harmony")

# Plot UMAP
DimPlot(
  obj.int,
  reduction = "umap.harmony",
  group.by = c("orig.ident"),
  combine = TRUE, label.size = 4
  )

# Save object
saveRDS(obj, "~/Documents/Dartmouth Class Work/Fall 2025/QBS145/obj.int.rds")
```

**3.	Cluster cells together using the integrated values. Set the resolution so you get 10 clusters (this is not what you should do on real data but makes grading easier). 10 points**
```{r}

```





